---
title: "Robust EZ-DDM explorations"
author: "Adriana, Eunice, and Joachim"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
    cards: false
    highlight: tango
    fig_width: 12 
    fig_height: 8 
    css: src/custom.css
---

```{r, echo = FALSE, message = FALSE}
library(here)
library(knitr)
library(kableExtra) 
library(tidyverse)
library(R2jags)

# Load custom functions
for(archive in dir(here("src"))){    source(paste(here("src/"),archive,sep=""))     }
```

# Introduction

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create the data frame
my_data <- data.frame(
  FactorB = c("EZ-DDM", "Robust EZ-DDM"),
  FactorA1 = c("Test 1", "Test 4"),
  FactorA2 = c("Test 2", "Test 5"),
  FactorA3 = c("Test 3", "Test 6")
)

# Generate the table using kable
kable(my_data, 
      col.names = c("Model","Clean trial-by-trial data", "Clean summary statistic data", "Trial-by-trial data with outliers"), # Set column names, empty for the first
      align = 'lccc') %>%                        # Align columns (left, center, center)
  kable_styling(bootstrap_options = c("striped", "bordered"), # Add styling
                full_width = FALSE)             # Prevent table from spanning full page width
```


## True parameter values

```{r}
# True hierarchical means
drift_mean <- 0      # Drift rate mean
bound_mean <- 1.5    # Boundary separation mean
nondt_mean <- 0.4    # Non-decision time mean

# Design settings
nParticipants <- 40
nTrials <- 160
```

```{r}
# Create a set of individual parameters
indiv_params <- sample_indivParams(n_participants = nParticipants, 
                                   dmean = drift_mean, bmean = bound_mean, nmean = nondt_mean)

# Print the first 10 rows showing the individual parameters
head(indiv_params, 10)
```


# Test 1 and 2: Clean trial-by-trial data

Tests 1 and 2 use clean data sampled from a Wiener process.

## Test 1: EZ-DDM

## Test 2: Robust EZ-DDM

# Test 3 and 4: Clean summary statistic data

Tests 3 and 4 use clean data (i.e., no outliers), but instead of generating trial-by-trial data, we use the EZ-DDM equations and implied sampling distribution to generate summary statistics.

## Test 3: EZ-DDM with outliers

## Test 4: Robust EZ-DDM with outliers

# Test 5 and 6: Trial-by-trial data with outliers

Tests 5 and 6 use trial-by-trial data, with outlier RTs.

## Test 5: EZ-DDM with outliers

## Test 6: Robust EZ-DDM with outliers

